Investigation of thermal reactionomes reveals stress tolerance vs. resistance in

### 2014-09-23 doc/ApTranscriptome_manuscript.Rmd
updates to ms

## Species differences in thermally-responsive genes

@@

@@

Specifically, 
...categories that were over-expressed relative to null...

@@

## Sensitivity of thermally-responsive genes to changes in temperature

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures. Genes turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

For **Intermediate** and **Bimodal** genes, genes had two inflection points so it was not possible to calculate a single temperature of gene activation. Instead, we calculated the variance of 1,000 random temperatures, weighted by expression at that temperature. For **Intermediate** genes, a greater variance would be indicative of less sensitivity to changes in temperature. For **Bimodal** genes, a greater variance would be indicative of a more sensitivity to changes in temperature (convex reaction norm). We found that **Intermediate** genes from *A. carolinensis* had a greater variance of expression than those from *A. picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`), indicating greater inertia of gene expression to changes in temperature. However, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`) for the **Bimodal** genes.

@@

We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed) by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, for **Low** genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for **High** genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.

>>>>

@@

## Thermal responses involve greater down-regulation in the warmer-climate species

@@

In contrast to expectations, more genes did not respond to high temperatures in *A. carolinensis*. In fact, the only differences were for **Intermediate** genes in *A. carolinensis*, which disproportionatly switched to **Low** or **Bimodal** expression in *A. picea*. Thus, *A. carolinensis*, the warmer-climate species, appears to rely upon shutting down gene expression to a greater extent than *A. picea*, which activates gene expression...

@@

## Greater inertia of gene expression response to heat stress in the warmer climate species

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures, including *Bimodal* genes. Genes turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** genes we compared the mean temperatures of downregulation of gene expression among species at both high (> 21°) and low (< 21°C) temperatures. Consistent with the results for up-regulated genes, we found that with increasing temperatures, the **Intermediate** genes were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`). Overall, these results indicate greater intertia of gene expression to changes in temperature for *A. carolinensis* than *A. picea*.

@@

We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed) by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. While finding that genes upregulated to thermal extremes

Contrary to expectations, for **Low** genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for **High** genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.
